---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


Analysis of Activity Monitoring Device Data 
===========================================

## Loading and preprocessing the data
The following R script will load the raw data activity.csv from the same working directory, prepare the data for analysis by R package dplyr, and convert the date column into Date class in R. 

```{r loadprep}
Activity <- read.csv("activity.csv")
str(Activity)

## Prepare the data frame to be analyzed with dplyr package.
suppressPackageStartupMessages(library(dplyr))
Activity <- tbl_df(Activity)

## Convert date to Date format in R
Activity$date <- as.Date(Activity$date, format = "%Y-%m-%d")
```


## What is mean total number of steps taken per day?
* The total number of steps taken per day during the monitored period is plotted in the histogram below. Its mean and median are calculated with the R codes below.

```{r sumperday}
## calculate the total number of steps taken per day
sum_day <- Activity %>%
        group_by(date) %>%
        summarise(sum_step_day = sum(steps, na.rm = TRUE))

## Plot histogram of the total number of steps taken each day
suppressPackageStartupMessages(library(ggplot2))
qplot(sum_day$sum_step_day, 
      grom="histogram", binwidth = 1000, 
      main = "Histogram of Total Number of Steps Taken Each Day \n  (2012-10-01 to 2012-11-30)",
      xlab = "Steps taken each day",
      ylab = "Count of Days",
      fill = I("steelblue"),
      col = I("blue"),
      alpha = I(.9))

##Calculate and report the mean and median of the total number of steps taken per day
mean_step_day <- as.integer(mean(sum_day$sum_step_day))
median_step_day <- median(sum_day$sum_step_day)
```

* The mean of total number of steps taken per day is `r mean_step_day` steps and the median of total number of steps taken per day is `r median_step_day` steps. 


## What is the average daily activity pattern?
* The average steps taken in specific 5-minute interval is plotted in the following time serie plot to demonstrate the average daily activity patten. 

```{r actpattern}
## calculate the average steps taken during 5-min interval across all days
avg_interval <- Activity %>%
        group_by(interval) %>%
        summarise(avg_step_interval = mean(steps, na.rm = TRUE))

## add a time sequence to the table for plotting purpose
avg_interval$time <- seq(ISOdatetime(2012,10,1,0,0,0), ISOdatetime(2012,10,1,23,55,0), by=(60*5))

plot(avg_step_interval ~ time, avg_interval, 
     type = "l", xlab = "Time Interval (min)", 
     ylab = "Average Steps Taken", 
     main ="Average Number of Steps Taken in 5-minutes Interval across All Days \n (2012-10-01 to 2012-11-30)" )

## identify the maximum of average steps taken and the 5-min interval it occurs
max_interval <- avg_interval[avg_interval$avg_step_interval == max(avg_interval$avg_step_interval), ]

```

* The maximum average steps taken is `r as.integer(max_interval$avg_step_interval)` steps during the `r max_interval$interval` 5-minutes interval, i.e. averagely maxium steps walking in 5 min interval is `r as.integer(max_interval$avg_step_interval)` steps between 8:35 am - 8:40 am.

## Imputing missing values
* The R script below counted the missing NA values in the original data and replaced them with the meidan of the corresponding 5-minute interval across all days measured. The data with filled NA values is then plotted in the histogram below. 

```{r missingvlaues}
## count number of missing values in the orignal data set
num_na <- nrow(Activity[Activity$steps == "NA", ])

## filling in the missing values with the median of the avarage steps for the same 5-min interval across all days. 
Activity_filled <- Activity %>%
        group_by(interval) %>%
        mutate(StepMedian = median(steps, na.rm = TRUE))

Activity_filled[is.na(Activity_filled$steps), ] <- mutate(Activity_filled[is.na(Activity_filled$steps), ], steps = StepMedian)

## plot the data with missing value filled into a histogram
## calculate the total number of steps taken per day
sum_day_filled <- Activity_filled %>%
        group_by(date) %>%
        summarise(sum_step_day = sum(steps))

qplot(sum_day_filled$sum_step_day, 
      grom="histogram", binwidth = 1000, 
      main = "Histogram of Total Number of Steps Taken Each Day \n (Missing Value Filled with Median of 5-min interval) \n  (2012-10-01 to 2012-11-30)",
      xlab = "Steps taken each day",
      ylab = "Count of Days",
      fill = I("steelblue"),
      col = I("blue"),
      alpha = I(.9))

##Calculate and report the mean and median of the total number of steps taken per day
mean_step_day_filled <- as.integer(mean(sum_day_filled$sum_step_day))
median_step_day_filled <- median(sum_day_filled$sum_step_day)


```

* There are `r num_na` missing values in the step counts. After filling these missing value with the median of corresponding 5-mintute interval across all days counted, the mean of total number of steps taken per day is `r mean_step_day_filled` and the median of total number of steps taken per day is `r median_step_day_filled`. 
* The histogram above and the mean after filling the missing values with the median of corresponding 5-minute interval are different from simply removing those NA values. The median stayed the same since the median of each interval was used to fill in NA values. The mean increased from `r mean_step_day` steps to `r mean_step_day_filled` steps since NA values were replaced thus the total steps and mean were increased. As shown in the histogram, the group with steps count between 5000 to 20000 were not changed much. But the days with 0 - 1000 step counts decreased from 10 days to 2 days while the groups with step counts between 1000 - 2000 increased from 0 to 8. Apparently, these were the days with most NA values which have been replaced with median of 5-min intervals. 

## Are there differences in activity patterns between weekdays and weekends?
* The follwoing R script plots the mean of steps taken in 5-minute intervals on weekdays v.s. on weekends to compare the activity patterns. 

```{r weekdiffweekend}

## add the column to desigate "weekday" or "weekend" levels
Weekday <- weekdays(Activity_filled$date)
Weekday[Weekday %in% c("Saturday", "Sunday")] <- "weekend"
Weekday[Weekday %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")] <- "weekday"
Activity_filled$weekvar <- as.factor(Weekday)

## Calculate mean of each time interval for both weekdays and weekends
Activity_week_mean_summary <- Activity_filled %>%
        group_by(weekvar, interval) %>%
        summarise(WeekMean = mean(steps))

## add a time sequence to the table for plotting purpose
## this is not the ideal way to create the x-axis since it still 
## contains the date value but sufficient for exploration purpose.
## I was able to convert the interval to POSIXlt format but the new 
## matrix keeps crashing R Studio when I try to use the new time 
## variable to plot. So, I used the alternative way below.   
Activity_week_mean_summary$time <- seq(ISOdatetime(2012,10,1,0,0,0), ISOdatetime(2012,10,2,23,55,0), by=(60*5))


## Plot the data into two panels to show activity patterns between weekdays and weekends
library(scales)
g <- ggplot (Activity_week_mean_summary, aes(time, WeekMean))
        g + geom_line() + facet_wrap(~ weekvar, nrow = 2, scales = "free_x") + 
                labs(y = "Average Steps Taken") + 
                labs(x = "5-minute Intervals ") + 
                labs(title = "Average Steps Taken across the Day during Weekdays and Weekends") +
        scale_x_datetime(labels = date_format("%H:%M"), breaks = date_breaks("2 hour"))

```

* Based on the two plots above, the maximus steps during 5-min intervals on weekdays is about 50 steps higher than weekends. During weekdays, the more active patterns (more than 100 steps per 5 minutes) concentrated between 8:00- 9:30 and 18:00 - 19:30; while during weekends, those 100+ steps/5min activities are more distributed throughout the day. 


